---
title: "midterm_dsII"
output: pdf_document
date: "2026-03-25"
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  echo = TRUE
)
```

```{r,echo=TRUE}
library(tidymodels)
library(caret)
library(pls)
library(earth)
library(mgcv)
library(splines)
library(corrplot)
library(car)
library(patchwork)
```

## 1. Load, clean and EDA analysis

Load, clean, and prepare dataset A for creating potential predictive models. 

```{r, message=FALSE, warning=FALSE}
load("data/datA.RData")
summary(datA)

# checking for any NAs in data 

colSums(is.na(datA)) 

# Select only numeric columns for the matrix
numeric_data = datA |> select(where(is.numeric))

# Calculate and plot
cor_matrix = cor(numeric_data, use = "complete.obs")
corrplot(cor_matrix, method = "circle", type = "upper")
```

There are a few correlated variables, which intuitively make sense:

  * `tumor_size`, `lymph_nodes`, and `stage` are highly positively correlated
  * `age` and `menopause` are positively correlated
  * `Ki67` (measures how fast cancer cells in a tumor are dividing) and `tumor_grade` (how abnormal cancer cells look) are positively correlated
  * `Ki67` and `ER` (estrogen receptor) are negatively correlated
  * `PR` (progesterone receptor) and `ER` are positively correlated
  * `hormonal_therapy` and `ER` are positively correlated


```{r}
# converting categorical variables into factors, as described in the assignment table 

datA = 
  datA |> 
  mutate(
    menopause = factor(menopause, levels = c("0", "1")),
    race = factor(race, levels = c("1", "2", "3", "4")),
    tumor_grade = factor(tumor_grade, levels = c("1", "2", "3")),
    stage = factor(stage, levels = c("1", "2", "3")),
    surgery = factor(surgery, levels = c("0", "1")),
    chemo = factor(chemo, levels = c("0", "1")),
    hormonal_therapy = factor(hormonal_therapy, levels = c("0", "1")),
    ER = factor(ER, levels = c("0", "1")),
    PR = factor(PR, levels = c("0", "1")),
    HER2 = factor(HER2, levels = c("0", "1"))
  ) |> 
  select(-id) # removing id since it isnt predictive 

# splitting data into training and testing for model training 

set.seed(1)
dat_split <- initial_split(datA, prop = 0.8)
trainData <- training(dat_split)
testData <- testing(dat_split)
```

# 2. Fitting a linear model
Calculating VIF on training data to see if the correlations observed earlier are multicollinear:
```{r}
# Fit a standard linear model first
vif_model = lm(log_PFS ~ ., data = trainData)

# Calculate VIF
vif(vif_model)
```

No variables are multicollinear, so all can be kept in models. 

Checking for interactions:
```{r}
ggplot(trainData, aes(x = tumor_size, y = log_PFS, color = stage)) +
    geom_point(alpha = 0.5) +
    geom_smooth(method = "lm", se = FALSE) 

ggplot(trainData, aes(x = age, y = log_PFS, color = menopause)) +
    geom_point(alpha = 0.5) +
    geom_smooth(method = "lm", se = FALSE) 
```
It looks like age/menopause and tumor size/cancer stage have an interaction. When we run the MARS model, we can take a closer look at these interactions. 

# 3. Fitting lm, knn, enet

We employed 10 fold cv

```{r}
# Using 10 fold cross validation for model comparison 
set.seed(1)
folds <- createFolds(trainData$log_PFS, k = 10)
ctrl <- trainControl(method = "repeatedcv",
                     number = 10,
                     repeats = 5,
                     index = folds)

# Linear regression

set.seed(1)
fit.lm <- train(log_PFS ~ .,
                data = trainData,
                method = "lm",
                trControl = ctrl)

# K-nearest neighbors

set.seed(1)
fit.knn <- train(log_PFS ~ .,
                 data = trainData,
                 method = "knn",
                 preProcess = c("center", "scale"),
                 tuneGrid = expand.grid(k = 1:25),
                 trControl = ctrl)

# Grabbing best params 
fit.knn$bestTune

# Elastic net 
set.seed(1)
fit.enet <- train(log_PFS ~ .,
                  data = trainData,
                  method = "glmnet",
                  tuneGrid = expand.grid(
                    alpha = seq(0, 1, length = 11),
                    lambda = exp(seq(-5, 2, length = 50))
                  ),
                  trControl = ctrl)

# Grabbing best params 
fit.enet$bestTune
```

# 4. Fitting pls, pcr, gam, mars 

```{r}
# Prepare matrices
x_train <- model.matrix(log_PFS ~ ., trainData)[, -1]
y_train <- trainData$log_PFS
x_test <- model.matrix(log_PFS ~ ., testData)[, -1]
y_test <- testData$log_PFS

set.seed(1)
pcr_model <- train(
  log_PFS ~ .,
  data = trainData,
  method = "pcr",
  preProcess = c("center", "scale"),
  trControl = ctrl,
  tuneLength = 20
)

set.seed(1)
pls_model <- train(
  log_PFS ~ .,
  data = trainData,
  method = "pls",
  preProcess = c("center", "scale"),
  trControl = ctrl,
  tuneLength = 20
)

# GAM
set.seed(1)
gam_model <- train(log_PFS ~ .,
                   data = trainData,
                   method = "gam",
                   trControl = ctrl)

# MARS
set.seed(1)
mars_model <- train(log_PFS ~ .,
                    data = trainData,
                    method = "earth",
                    tuneGrid = expand.grid(degree = 1:3, nprune = 2:15),
                    trControl = ctrl)
```

# 5. Comparing all models training RMSE

```{r}
all_models <- resamples(list(
  LM = fit.lm,
  KNN = fit.knn,
  ENET = fit.enet,
  PCR = pcr_model,
  PLS = pls_model,
  GAM = gam_model,
  MARS = mars_model
))

# Extract just the RMSE statistics matrix
rmse_matrix <- summary(all_models, metric = "RMSE")$statistics$RMSE
cv_means <- rmse_matrix[, "Mean"]
best_model_name <- names(cv_means)[which.min(cv_means)]
best_model_rmse <- min(cv_means)
worst_model_name <- names(cv_means)[which.max(cv_means)]
worst_model_rmse <- max(cv_means)

rmse_matrix |> knitr::kable(digits = 4)
bwplot(all_models, metric = "RMSE")
cv_rmse = summary(all_models)$statistics$RMSE[, "Mean"]

# evaluating test predictions

predictions <- list(
  LM = predict(fit.lm, testData),
  KNN = predict(fit.knn, testData),
  ENET = predict(fit.enet, testData),
  PCR = predict(pcr_model, testData),
  PLS = predict(pls_model, testData),
  MARS = predict(mars_model, testData),
  GAM = predict(gam_model, testData)
)

# Post-resample RMSE on test set
test_rmse <- sapply(predictions, function(pred) postResample(pred, testData$log_PFS)[["RMSE"]])

# summarize RMSE in table
model_comparison = tibble(
  model = names(test_rmse),
  cv_rmse = cv_rmse,
  test_rmse = test_rmse
  ) |> 
  arrange(test_rmse) 
model_comparison |> knitr::kable(digits = 3)
```

Final Coefficients: 
```{r}
enet_coefs = coef(fit.enet$finalModel, fit.enet$bestTune$lambda) |> as.matrix()

enet_coefs_clean = tibble(
  predictor = rownames(enet_coefs),
  coefficient = enet_coefs[, 1]
  ) |> 
  arrange(desc(abs(coefficient)))

enet_coefs_clean |> knitr::kable(digits = 3)
```

Warning: Some resampling iterations for the elastic net model produced missing performance values, likely due to numerical instability at extreme tuning parameter values. However, the majority of resamples were valid, and the selected tuning parameters were stable.

We used repeated 10-fold cross-validation to estimate out-of-sample prediction error. This approach repeatedly partitions the training data into training and validation sets, reducing variance compared to a single split. We evaluated seven candidate models: Linear Regression, K-Nearest Neighbors, Elastic Net (ENET), Principal Component Regression (PCR), Partial Least Squares (PLS), Generalized Additive Models (GAM), and Multivariate Adaptive Regression Splines (MARS).

Based on the training data, most models performed very similarly, with mean RMSE values ranging from `r round(best_model_rmse, 4)` (`r best_model_name`) to `r round(worst_model_rmse, 4)` (`r worst_model_name`). The similar performance across linear and nonlinear models suggests that the relationship between the predictors and `log(PFS)` is largely linear.

PLS achieved the lowest mean error, closely followed by PCR (Mean = `r round(cv_means[4], 4)`) and ENET (Mean = `r round(cv_means[3], 4)`). ENET and PCR performing similarly to PLS suggests that dimensionality reduction provided little additional predictive power. ENET has the lowest maximum RMSE observed. With the exception of one outlier, the remaining RMSEs from cross-validation have a tight distribution, indicating this model is less variable and more stable. 

Because predictive accuracy was comparable across the top models, we prioritized interpretability. We selected the ENET model, which was trained using an explicit lambda grid to ensure consistent tuning across folds. The cross-validation process selected an alpha of 0, meaning the final model is a pure Ridge regression. While Ridge regression does not perform sparse feature selection (all variables are retained), it stabilizes the coefficients to handle multicollinearity, making it a highly robust and easily interpretable model for our clinical context. The reported test error for this ENET model was `r round(unname(test_rmse[3]), 4)`. 

# 6. Load, clean, and prepare dataset B
```{r}
load("data/datB.RData")
summary(datB)

# checking for any NAs in data 

colSums(is.na(datB)) 

# Select only numeric columns for the matrix
numeric_data_B = datB |> select(where(is.numeric))

# Calculate and plot
cor_matrix_B = cor(numeric_data_B, use = "complete.obs")
corrplot(cor_matrix_B, method = "circle", type = "upper")
```

Correlations are similar to datA. Proceeding with same data cleaning
```{r}
# converting categorical variables into factors, as described in the assignment table 

datB = 
  datB |> 
  mutate(
    menopause = factor(menopause, levels = c("0", "1")),
    race = factor(race, levels = c("1", "2", "3", "4")),
    tumor_grade = factor(tumor_grade, levels = c("1", "2", "3")),
    stage = factor(stage, levels = c("1", "2", "3")),
    surgery = factor(surgery, levels = c("0", "1")),
    chemo = factor(chemo, levels = c("0", "1")),
    hormonal_therapy = factor(hormonal_therapy, levels = c("0", "1")),
    ER = factor(ER, levels = c("0", "1")),
    PR = factor(PR, levels = c("0", "1")),
    HER2 = factor(HER2, levels = c("0", "1"))
  ) |> 
  select(-id) # removing id since it isnt predictive
```

# 7. Testing ENET model on dataset B

```{r}
enet_pred_B = predict(fit.enet, newdata = datB)
enet_test_rmse <- RMSE(pred = enet_pred_B, obs = datB$log_PFS)
print(enet_test_rmse)
```

When applying the final Elastic Net model (trained on datA) to the new cohort (datB), the RMSE increased to `r round(enet_test_rmse, 4)`. Compared to the previous test RMSE of `r round(unname(test_rmse[3]), 4)` on the original data, this indicates a drop in predictive performance. This gap suggests that there may be underlying distribution shifts between the two patient populations (e.g., differences in demographics, clinical settings, or unmeasured variables), or that the model slightly overfit the original cohort despite the use of Ridge regularization. A new model will be needed. Since ENET had a tight distribution when using the original data, it seems the most promising (rather than training many different models). Additionally, it is then easier to compare what changed in the model. 

# 8. Trainning new model for dataset B

```{r}
# splitting data into training and testing for model training 

set.seed(1)
dat_split <- initial_split(datB, prop = 0.8)
trainData <- training(dat_split)
testData <- testing(dat_split)

# Elastic net 
set.seed(1)
fit.enet.new <- train(log_PFS ~ .,
                      data = trainData,
                      method = "glmnet",
                      tuneGrid = expand.grid(
                        alpha = seq(0, 1, length = 11),
                        lambda = exp(seq(-5, 5, length = 50))
                        ),
                      trControl = ctrl)

# Grabbing best params 
fit.enet.new$bestTune

enet_new_pred = predict(fit.enet.new, newdata = testData)
enet_new_test_rmse <- RMSE(pred = enet_new_pred, obs = testData$log_PFS)
print(enet_new_test_rmse)
```

This RMSE looks more similar to the test RMSE from the original data. 

Comparing the old and new model:
```{r}
enet_old_coefs = as.matrix(coef(fit.enet$finalModel, fit.enet$bestTune$lambda))

coefs_old = tibble(
  predictor = rownames(enet_old_coefs),
  coefficient = enet_old_coefs[, 1]) |> 
  arrange(desc(abs(coefficient))) 

coefs_old |> knitr::kable(digits = 4)

enet_new_coefs = as.matrix(coef(fit.enet.new$finalModel, fit.enet.new$bestTune$lambda))

coefs_new = tibble(
  predictor = rownames(enet_new_coefs),
  coefficient = enet_new_coefs[, 1]) |> 
  arrange(desc(abs(coefficient))) 

coefs_new |> knitr::kable(digits = 4)

merged_coefs = 
  full_join(coefs_old, coefs_new, 
            by = "predictor", suffix = c("_old", "_new")) |> 
  mutate(difference = coefficient_old - coefficient_new)

merged_coefs |> arrange(predictor)  |> knitr::kable(digits = 4)
```

By comparing the coefficients of the old pre-2000 model to the new 2000-2005 model, several major predictive shifts become clear. The new model's intercept increased from 2.55 to 3.03, indicating that the baseline PFS increased over the decade. This makes sense since new and better interventions will have likely been more accessible. The impact of HER2 status (`HER21`) has flipped. In the older cohort, being HER2+ was a severe negative predictor for progression-free survival (-0.101). In the new cohort, this penalty essentially vanished (+0.009). Other predictors that flipped were all race categories, `menopause1`, `stage2`, and `tumor_grade2`. Others like ER and hormonal therapy stayed largely the same and continued to improve PFS, while stage 3, tumor grade 3, and tumor size decreased PFS. The gap in predictive power makes sense given these changes in cohort trends. 
